---
title: " More `ggplot` and intro to the `library(dplyr)` "
author: 'Víctor Peña'
output:
  html_document:
    toc: true
---

## More `qplot`

```{r}
library(ggplot2)
data(diamonds)
```

### More than 2 variables at once

We can create scatterplots with colored dots of different shapes:

```{r, message = FALSE}
qplot(x=carat, y=price, color=cut, shape=cut, data=diamonds)
```

We can plot colored smoothed curves (potentially overlaid on points; not recommended, though).

```{r}
qplot(x=carat, y=price, color=cut, geom='smooth', data=diamonds)+geom_point(alpha=0.02)
```

We can plot panels with colored dots:
```{r}
qplot(x=carat, y=price, color=cut, facets=color ~ ., data=diamonds)
```

Double panels!
```{r}
qplot(x=carat, y=price, facets=color~cut, data=diamonds)
```

Double panels of colored dots... [Q: how many variables are we plotting at once now?]
```{r}
qplot(x=carat, y=price, color=clarity, facets=color~cut, data=diamonds)
```

We could redo all of these with smooth curves/lines instead. 

How can we plot relationships between 3 numerical/quantitative variables? This is one option

```{r}
qplot(x=depth, y=carat, color=price, data=diamonds)
```

Another option is categorizing one of the variables, and then plotting bivariate relationships in panels. Below we partition the variable `depth` into 4 categories (defined by its quartiles):

```{r}
diamonds$depthcat = cut(diamonds$depth, breaks=quantile(diamonds$depth), include.lowest = TRUE)
qplot(x=carat, y=price, facets=.~depthcat, data=diamonds)

```

### `grid.arrange`

Sometimes it's useful to have unrelated plots in one panel. 

**Exercise** With the base `graphics` package, we can do that with `par(mfrow=c( , ))`. For example, if we want a plot that has 2 rows, one with a histogram of `price` and another row which has a bar plot of cut, how would we do that?

**Answer**

```{r}
par(mfrow=c(2,1))
hist(diamonds$price)
plot(diamonds$cut)
```

Unfortunately, `par(mfrow=c(,))` doesn't work with `ggplot`. Fortunately, we have `grid.arrange` in `library(gridExtra)`:

```{r}
library(gridExtra)
p1 = qplot(price, data=diamonds)
p2 = qplot(cut, data = diamonds)
grid.arrange(p1, p2, nrow=2)
```

If you want to learn more options, you can go [here](https://cran.r-project.org/web/packages/egg/vignettes/Ecosystem.html).

### Some cool plots in `library(GGally)`

There are some good extra plots based on `ggplot2` in `library(GGally)`. One of them is the equivalent of `plot(dataset)`:
```{r, message=FALSE}
library(GGally)
ggpairs(diamonds[,5:10])
```

You can plot confidence intervals for regression coefficients easily:
```{r, message=FALSE}
data(mtcars)
mod = lm(mpg~wt+qsec+cyl,data=mtcars)
ggcoef(mod)
```

## More `ggplot`
 
Today we'll cover more `ggplot`. Last time we used `qplot` and today we'll use the more general `ggplot` machinery. Plotting with `ggplot` can be more or less broken down into the following steps.

1. Read in the data 

2. Add *aesthetics* (`aes`): Which variables go into the plot? What are the `x` and `y`? What variables are you using for color-coding?

3. Add `geom`s: What kind of plot do you want?

4. Do data transformations, if needed.

5. Change labels/theme.

We'll see some concrete examples today.

Let's read in the diamonds data first.

```{r}
library(ggplot2)
data(diamonds)
```

Let's create a stacked % barplot:

```{r}
ggplot(diamonds) + aes(x=cut, fill=color) + geom_bar()
```

We can change the type of barplot easily by adding options in the `geom_bar` function:

```{r}
ggplot(diamonds) + aes(x=cut, fill=color) + geom_bar(position='fill')
```

If you want side-by-side bars, you can use `position = 'dodge' `

```{r}
ggplot(diamonds) + aes(x=cut, fill=color) + geom_bar(position='dodge')
```

We can use the same structure to produce a color-coded scatterplot:

```{r}
p = ggplot(diamonds) + aes(x=carat, y=price, color=cut) + geom_point()
p
```

Note that we used the option `color` in `aes`. 

We can break down plots in panels using `facet_grid`:

```{r}
p = p + facet_grid(. ~ cut)
p
```

And you can combine `geom`s. For example

```{r}
ggplot(diamonds) + aes(x=price) + geom_histogram(binwidth = 500, aes(y=..density..))+geom_density(color='red', size=1)
```

## Axes, titles and labels

The relevant commands here are

* `ggtitle`: for changing the title

* `xlab`, `ylab`: $x$ and $y$ labels

* `xlim`, `ylim`: limits / scale of the plot


```{r}
p + xlab("carat (weight)") + ylab("price ($)") + ggtitle("Price vs carat vs quality") + xlim(c(0,10))+ylim(c(0,30000))
```

You can also transform the axes


```{r}
p + coord_trans(x='log',y='log')
```


## Intro to the `library(dplyr)`

We can manipulate / filter datasets easily with functions in `library(dplyr)`. Let's use the `hsb2` dataset in `library(openintro)` to illustrate some of the functions.

```{r, message=FALSE}
library(dplyr)
library(openintro)
data(hsb2)
hsb2 = as_tibble(hsb2)
```

### `select`: select / drop variables

You can create subsets of the data that only contain a few of the variables with `select`. For example, if you want to create a subset that only has the variables `math`, `race`, `gender`, and `ses`:

```{r}
sub1 = hsb2 %>% select(-math,-race,-gender,-ses)
sub1
```

The command `%>%` is the so-called pipe operator, which can be used to combine functions in `dplyr` (it's kind of the same idea as `+` in `library(ggplot2)`). 

If, on the other hand, you want to create a subset that **excludes** the variables `math`, `race`, `gender`, and `ses`:

```{r}
sub2 =  hsb2 %>% select(-math,-race,-gender,-ses)
sub2
```


### `filter`: filter by logical conditions

This one is pretty self-explanatory. As a reminder, the logical operators in `r` are:

* `==`: equal to

* `!=`: not equal to

* `>=`, `>`, `<`, `<=`: greater or equal to, greater than, less than, less than or equal to

* `|`: or

* `&`: and

For example, if you want to create a subset that only contains people who went to public school and got a score in `math` greater than 70, you can do that as follows:

```{r}
sub3 = hsb2 %>% filter(math > 70, schtyp == 'public') 
sub3
```

If, on the other hand, you want to filter those who got a score in `math` greater than 70 or went to public school (or both):

```{r}
hsb2 %>% filter(math > 70 | schtyp == 'public') 
```

You can combine `select` and `filter`. For example, if you only want to keep the values of `math` and `schtyp` for people who went to public school and got a score in `math` greater than 70, we can do that as follows

```{r}
sub3 = sub3 %>% select(math, schtyp)
```

Or, equivalently

```{r}
sub3 =  hsb2 %>% filter(math > 70, schtyp == 'public') %>% select(math, schtyp)
```

### `mutate`: transform variables

We can use mutate if we want to transform/create new variables. For example, if we want to create a new variable called `avg` which contains the average schore in `read`, `write`, `science`, and `socst`:

```{r}
hsb2 =  hsb2 %>% mutate(avg=(read+write+science+socst)/4)
```


### `arrange`: sort 

You can use `arrange` to sort the data by the values of some variable (default is ascending; you can use `desc()` to sort in descending order). 

For example, if you want to sort by `avg`, which is the new variable we created in the previous section:

```{r}
hsb2 %>% arrange(avg)
```

If you want to sort in descending order

```{r}
hsb2 %>% arrange(desc(avg))
```


### `group_by` and `summarize`: obtain summaries by variables

We can create objects which contain summaries for different groups by combining `group_by` and `summarize`:

```{r}
hsb2 %>% group_by(race) %>% summarize(medMath = median(math), sdMath = sd(math))
```

And you can combine these function with the other functions we learned today. For example:

```{r}
hsb2 %>% group_by(race) %>% filter(math > 70) %>% summarize(n=n())
```

Tells us that there are 10 people who got a `math` score greater than 70, and that 1 of them is `asian` and 9 of them are `white`. 

### More?

* Next time

* [`dplyr` cheat sheet](http://www.rstudio.com/wp-content/uploads/2015/02/data-wrangling-cheatsheet.pdf)
