---
title: "Combining datasets "
author: "Víctor Peña"
output:
  html_document:
    toc: true
    theme: cosmo
---

```{r, echo=FALSE, message=FALSE, warning=FALSE}
library(dplyr)
library(tidyr)
```


## Joining data 

In lecture, we learned about `left_join`, `right_join`, `inner_join`, and `full_join`. Our starting point is 2 datasets  `a` and `b`, which have a variable in common named `x`. We wish to *merge* the datasets by matching values of `x`. 

* `left_join(a, b, by='x')`: Keep all the rows in `a`, find rows in dataset `b` that have matching values of `x` in `a`, and include the extra columns in `b` that don't appear in `a`. Don't include rows in `b` that have no matching value of `x` in `b`.

* `right_join(a, b, by='x')`: Keep all the rows in `b`, find rows in dataset `a` that have matching values of `x` in `b`, and include the extra columns in `a` that don't appear in `b`. Don't include rows in `a` that have no matching value of `x` in `b`.

* `inner_join(a, b, by='x')`: Keep rows that have values of `x` that appear in **both** `a` and `b`.

* `full_join(a, b, by='x')`: Keep all the rows in `a` and `b`, even if they don't have matching values in `x`.


To make this concrete, let's work with a specific example. The `tibbles` below contain names of drinks:

```{r}
nonalcoholic = tibble(name=c("g&t","mimosa","rum and coke","calimocho","polar"), 
                      mixer=c("tonic","orange juice", "coke", "coke","seltzer"))
alcoholic = tibble(name=c("g&t","mimosa","rum and coke","calimocho","IPA","dark and stormy"), 
                    spirit=c("gin","sparkling wine", "rum", "red wine","beer","rum"))
```

Let's try out what happens when we try the different `join`s on the `tibble`s:

```{r}
left_join(nonalcoholic, alcoholic, by='name') 
```

```{r}
right_join(nonalcoholic, alcoholic, by='name') 
```

```{r}
inner_join(nonalcoholic, alcoholic, by='name') 
```

```{r}
full_join(nonalcoholic, alcoholic, by='name')
```

We can also use `join`s for filtering. For example, `semi_join(a, b, by='x')` keeps all the rows in `a` that have a matching value of `x` in `b` (it doesn't create more columns with the information in `b`), and `anti_join(a, b, by='x')` keeps the rows in `a` that don't have a matching value of `x` in `b`:

```{r}
semi_join(nonalcoholic, alcoholic, by='name')
```

```{r}
anti_join(nonalcoholic, alcoholic, by='name')
```

Suppose we get prices for the drinks:

```{r}
nonalcoholic = nonalcoholic %>% mutate(price=c(8, 6, 7, 5, 5))
alcoholic = alcoholic  %>% mutate('$'=c(8, 6, 7, 5, 5, NA))
```

How can we join the new `tibbles` nicely?

First, we can rename the price variable in `alcoholic`:

```{r}
alcoholic = alcoholic %>% rename(price='$')
```

And then, we can `full_join` the `tibbles` by their `name` and `price` (we want a double match):

```{r}
join = full_join(nonalcoholic, alcoholic, by=c('name','price'))
join
```

If we want to reorder the columns so that the spirit comes first, followed by the mixer, we can use `select`:

```{r}
join = join %>% select(name, spirit, mixer, price)
join
```

Now, suppose we got a new `tibble` with the extra components that were missing for `dark and stormy`

```{r}
ds = tibble(name='dark and stormy', spirit='ginger beer', price = 8)
```

How can we incorporate this new information?

We can start by `full_join`ing

```{r}
full_join(join, ds, by='name')
```

This `join` has a few problems. We'd like to combine `price.x` and `price.y`. We can do that using the function `coalesce`, and then getting rid of `price.x` and `price.y`:

```{r}
full_join(join, ds, by='name') %>%  
  mutate(price=coalesce(price.x,price.y)) %>% select(-price.x, -price.y)
```

This is mostly fine. If we want to combine `spirit.x` and `spirit.y`:

```{r}
full_join(join, ds, by='name') %>%  
  mutate(price=coalesce(price.x,price.y)) %>% select(-price.x, -price.y) %>%
    replace(is.na(.),"") %>%
    unite(spirit, spirit.x, spirit.y, sep="")
```

First, we replaced the `NAs` by empty spaces, and then we joined the variables `spirit.x`, and `spirit.y` using `unite`. The formatting isn't great, unforunately. We can play around a little bit to make it look nicer: 

```{r}
ds = ds %>% mutate(spirit=paste(" ", spirit,sep=""))
final = full_join(join, ds, by='name') %>%  
  mutate(price=coalesce(price.x,price.y)) %>%
  select(-price.x, -price.y)  %>%
  replace(is.na(.),"") %>%
  unite(spirit, spirit.x, spirit.y, sep="")
final
```

Now suppose that we're at some bar where the total price for a mixed drink is equal to the sum of the prices of the components. The prices are

```{r}
nonalcoholic = tibble(name=c("g&t","mimosa","rum and coke","calimocho","polar"), 
                      mixer=c("tonic","orange juice", "coke", "coke","seltzer"))  %>% 
                      mutate(price=c(3, 2, 2, 2, 5))
alcoholic = tibble(name=c("g&t","mimosa","rum and coke","calimocho","IPA"), 
                    spirit=c("gin","sparkling wine", "rum", "red wine","beer")) %>% 
                    mutate(price=c(5, 4, 5, 3, 5))

```

How can we merge the data and combine (add) prices? 

We can start by `full_join`ing by name

```{r}
full_join(nonalcoholic, alcoholic, by='name')
```

Unfortunately, a simple `mutate` instruction isn't satisfactory because the sum of an `NA` + a number is equal to `NA`:

```{r}
full_join(nonalcoholic, alcoholic, by='name') %>% mutate(total=price.x+price.y)
```

We can solve this problem in different ways. One way is using `coalesce` to replace the `NA`s by zeros in `price.x` and `price.y`, and then adding:

```{r}
full_join(nonalcoholic, alcoholic, by='name') %>% 
  mutate(price.x=coalesce(price.x,0), price.y=coalesce(price.y,0), total=price.x+price.y) %>% 
  select(name, spirit, mixer, total)
```

In general, the way that `R` deals with `NA`s can be a little frustrating, as these examples show. 

## Set operations and binding 

Source: [Data wrangling cheat sheet](https://www.rstudio.com/wp-content/uploads/2015/02/data-wrangling-cheatsheet.pdf)

```{r, echo = FALSE}
y = tibble(x1 = c("A", "B", "C"), x2 = 1:3)
z = tibble(x1 = c("B", "C", "D"), x2 = 2:4)
```

Consider the `tibble`s `y` and `z`, which have matching variable names `x1` and `x2` (this is important!):
```{r}
y
z
```

**Set operations**

* `intersect(y,z)`: find rows that appear in both `y` and `z`

```{r}
intersect(y,z)
```

* `union(y,z)`: find rows that appear in `y` or `z` (or both):

```{r}
union(y,z)
```

* `setdiff(y, z)`: find rows that appear in `y` but not in `z`:

```{r}
setdiff(y,z)
```

**Binding rows and columns**

If you want to bind the rows of `y` and `z`
```{r}
bind_rows(y,z)
```

If you want to bind the columns of `y` and `z`:

```{r}
bind_cols(y,z)
```